\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2023-04-13 submission/manuscript.Rmd
minor updates

keywords          : "attractiveness, AI-generated images, fiction, fake news, sense of reality"
wordcount         : "5223"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies or via recent algorithms such as GANs or diffusion models often include distortions or lack certain key details that makes them visually distinct from real images, @mcdonnell2010face; @corvi2022detection], it is fair to assume that these technical limitations will become negligible in the near future. This is particularly true in the field of face generation, where face-generation algorithms are already able to create stimuli that are virtually indistinguishable from real photos [@nightingale2022; @tucciarelli2020; @moshel2022]. Such technological feat, however, leads to a new question: if real and fake stimuli cannot be differentiated based on their objective "physical" characteristics, how can we form judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2018cognitive; @makowski2019phenomenal]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus (e.g., which journal was the information published in), and its credibility, authority and expertise, to help facilitate their evaluation [@petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) makes this task increasingly difficult [@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial role in simulation monitoring, with factors such as cognitive style, prior beliefs, and personality traits [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with stronger analytical reasoning have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy], and prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information that is in line with one's expectations [@britt2019reasoned]. Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

@@

Another issue is the impact on reality judgements of the prior explicit instruction that "about half of the faces were AI-generated and the other half real photos". Given this prior information given to participants, it migjt seems like our enthusiasm pertaining to the finding that most people did indeed believe a high number of stimuli to be fake might be unwarranted: all it shows is that participants followed the instructions. First of all, even if that was the case, the fact that our beliefs of reality can be so easily re-programmed with basic instructions ad lead to high-confidence answers remains an interesting phenomenon. Moreover, it is to note that the paradigm did not instruct participants to balance their answers according to a certain distribution (e.g., 50-50), merely providing a description of the dataset. The fact that no presentation order effect was found on reality beliefs gives credit to the assumption that participants did not try to actively distribute their responses to match the instructions, in which case we would have expected a different pattern, for instance the first items judged as real (the initial "true" belief of the participants), and later a hard bias towards responding "fake" (as participants realize that all stimuli are of similar nature and that they have to "make up" for the prevalence of their "real" answers to fulfil the instructions).

That said, the potential confounding effect of the instructions still exists, and a control condition without the cover story with AI-generated images would in-principle be welcome. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum (asking them to discriminate real from fake without introducing the context and meaning of "fake" seems hardly feasible). That said, future studies should study the impact of these higher-order expectations on ratings [for instance, @tucciarelli2020 found that merely mentioning that some faces were AI generated decreased on average the trustworthiness ratings for all faces] as well as on the simulation monitoring process itself (i.e., the "criterion": would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., "most of the images *but a few* are real" vs. "most of the images *but a few* are fake") or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs. real to participants.

>>>>

keywords          : "attractiveness, AI-generated images, fiction, fake news, sense of reality"
wordcount         : "5242"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies or via recent algorithms such as GANs or diffusion models often include distortions or lack certain key details that makes them visually distinct from real images, @mcdonnell2010face; @corvi2022detection], it is fair to assume that these technical limitations will become negligible in the near future. This is particularly true in the field of face generation, where face-generation algorithms are already able to create stimuli that are virtually indistinguishable from real photos [@nightingale2022; @tucciarelli2020; @moshel2022]. Such a technological feat, however, leads to a new question: if real and fake stimuli cannot be differentiated based on their objective "physical" characteristics, how can we form judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2018cognitive; @makowski2019phenomenal]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus (e.g., which journal was the information published in), and its credibility, authority and expertise, to help facilitate their evaluation [@petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are often mass-shared with little context) makes this task progressively difficult [@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial role in simulation monitoring, with factors such as cognitive style, prior beliefs, and personality traits [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with stronger analytical reasoning skills have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy], and prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information that is in line with one's expectations [@britt2019reasoned]. Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

@@

Another issue is the impact on reality judgements of the prior explicit instruction that "about half of the faces were AI-generated and the other half real photos". Given this prior information given to participants, it might seems like our enthusiasm pertaining to the finding that most people did indeed believe a high number of stimuli to be fake might be unwarranted, since it simply affirms participants followed the instructions. However, even if that was the case, the finding that our beliefs of reality can be so easily re-programmed with simple instructions and lead to high-confidence answers remains an interesting phenomenon. Moreover, it is to note that the paradigm did not instruct participants to balance their answers according to a certain distribution (e.g., 50-50), merely providing them a description of the dataset. The fact that no presentation order effect was found on reality beliefs suggests that participants did not try to actively distribute their responses to match the instructions, in which case we would have expected a different pattern, for instance the first few items judged as real (the initial "true" belief of the participants), and progressively a hard bias towards responding "fake" (as participants realize that all stimuli are of similar nature and that they have to "make up" for the prevalence of their "real" answers to fulfill the expected proportion of responses given the instructions).

That said, the potential demand effect of the instructions still exists, and a control condition without the cover story with AI-generated images would in-principle be able to mitigate such confounds to some extent. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum (simply instructing them to discriminate real from fake without providing some background information regarding the context and defining what is meant by "fake" seems hardly feasible). That being the case, future studies should study the impact of these higher-order expectations on ratings [for instance, @tucciarelli2020 found that merely mentioning that some faces were AI generated decreased, on average, the trustworthiness ratings for all faces] as well as on the simulation monitoring process itself (i.e., the "criterion": would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., "most of the images *but a few* are real" vs. "most of the images *but a few* are fake") or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs. real to participants.

### 2023-04-13 submission/manuscript.tex
minor updates

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{attractiveness, AI-generated images, fiction, fake news, sense of reality\newline\indent Word count: 5223}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage[titles]{tocloft}
\cftpagenumbersoff{figure}
\renewcommand{\cftfigpresnum}{\itshape\figurename\enspace}
\renewcommand{\cftfigaftersnum}{.\space}
\setlength{\cftfigindent}{0pt}
\setlength{\cftafterloftitleskip}{0pt}
\settowidth{\cftfignumwidth}{Figure 10.\qquad}
\usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated},
  pdfauthor={Dominique Makowski1, An Shu Te1, Stephanie Kirk1, Ngoi Zi Liang1, Panagiotis Mavros2, \& S.H. Annabel Chen1, 3, 4, 5},
  pdflang={en-EN},
  pdfkeywords={attractiveness, AI-generated images, fiction, fake news, sense of reality},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

While not all simulations have achieved perfect realism (Corvi et al., 2022; e.g., Computer Generated Images - CGI in movies or via recent algorithms such as GANs or diffusion models often include distortions or lack certain key details that makes them visually distinct from real images, McDonnell \& Breidt, 2010), it is fair to assume that these technical limitations will become negligible in the near future. This is particularly true in the field of face generation, where face-generation algorithms are already able to create stimuli that are virtually indistinguishable from real photos (Moshel et al., 2022; Nightingale \& Farid, 2022; Tucciarelli et al., 2020). Such technological feat, however, leads to a new question: if real and fake stimuli cannot be differentiated based on their objective ``physical'' characteristics, how can we form judgements regarding their nature?

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality (a process henceforth referred to as \emph{simulation monitoring}, Makowski, 2018; Makowski, Sperduti, et al., 2019). With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy (Michael \& Sanson, 2021; Susmann et al., 2021), people tend to draw on peripheral contextual cues (\textbf{Figure 1}), such as the source of the stimulus (e.g., which journal was the information published in), and its credibility, authority and expertise, to help facilitate their evaluation (Michael \& Sanson, 2021; Petty \& Cacioppo, 1986; Susmann et al., 2021). However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) makes this task increasingly difficult (Berghel, 2018; Y. Chen et al., 2015). Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial role in simulation monitoring, with factors such as cognitive style, prior beliefs, and personality traits (Bryanov \& Vziatysheva, 2021; Ecker et al., 2022; Sindermann et al., 2020). For instance, individuals with stronger analytical reasoning have been found to better discriminate real from fake stimuli (Pehlivanoglu et al., 2021; Pennycook \& Rand, 2019), and prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information that is in line with one's expectations (Britt et al., 2019). Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news (Piksa et al., 2022; Sindermann et al., 2020).

@@

Another issue is the impact on reality judgements of the prior explicit instruction that ``about half of the faces were AI-generated and the other half real photos''. Given this prior information given to participants, it migjt seems like our enthusiasm pertaining to the finding that most people did indeed believe a high number of stimuli to be fake might be unwarranted: all it shows is that participants followed the instructions. First of all, even if that was the case, the fact that our beliefs of reality can be so easily re-programmed with basic instructions ad lead to high-confidence answers remains an interesting phenomenon. Moreover, it is to note that the paradigm did not instruct participants to balance their answers according to a certain distribution (e.g., 50-50), merely providing a description of the dataset. The fact that no presentation order effect was found on reality beliefs gives credit to the assumption that participants did not try to actively distribute their responses to match the instructions, in which case we would have expected a different pattern, for instance the first items judged as real (the initial ``true'' belief of the participants), and later a hard bias towards responding ``fake'' (as participants realize that all stimuli are of similar nature and that they have to ``make up'' for the prevalence of their ``real'' answers to fulfil the instructions).

That said, the potential confounding effect of the instructions still exists, and a control condition without the cover story with AI-generated images would in-principle be welcome. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum (asking them to discriminate real from fake without introducing the context and meaning of ``fake'' seems hardly feasible). That said, future studies should study the impact of these higher-order expectations on ratings (for instance, Tucciarelli et al., 2020 found that merely mentioning that some faces were AI generated decreased on average the trustworthiness ratings for all faces) as well as on the simulation monitoring process itself (i.e., the ``criterion'': would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., ``most of the images \emph{but a few} are real'' vs.~``most of the images \emph{but a few} are fake'') or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs.~real to participants.

@@

\leavevmode\vadjust pre{\hypertarget{ref-makowski2019phenomenal}{}}%
Makowski, D., Sperduti, M., Pelletier, J., Blondé, P., La Corte, V., Arcangeli, M., Zalla, T., Lemaire, S., Dokic, J., Nicolas, S., et al. (2019). Phenomenal, bodily and brain correlates of fictional reappraisal as an implicit emotion regulation strategy. \emph{Cognitive, Affective, \& Behavioral Neuroscience}, \emph{19}(4), 877--897.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rhodes2006evolutionary}{}}%
Rhodes, G. et al. (2006). The evolutionary psychology of facial beauty. \emph{Annual Review of Psychology}, \emph{57}, 199.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{attractiveness, AI-generated images, fiction, fake news, sense of reality\newline\indent Word count: 5242}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage[titles]{tocloft}
\cftpagenumbersoff{figure}
\renewcommand{\cftfigpresnum}{\itshape\figurename\enspace}
\renewcommand{\cftfigaftersnum}{.\space}
\setlength{\cftfigindent}{0pt}
\setlength{\cftafterloftitleskip}{0pt}
\settowidth{\cftfignumwidth}{Figure 10.\qquad}
\usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

While not all simulations have achieved perfect realism (Corvi et al., 2022; e.g., Computer Generated Images - CGI in movies or via recent algorithms such as GANs or diffusion models often include distortions or lack certain key details that makes them visually distinct from real images, McDonnell \& Breidt, 2010), it is fair to assume that these technical limitations will become negligible in the near future. This is particularly true in the field of face generation, where face-generation algorithms are already able to create stimuli that are virtually indistinguishable from real photos (Moshel et al., 2022; Nightingale \& Farid, 2022; Tucciarelli et al., 2020). Such a technological feat, however, leads to a new question: if real and fake stimuli cannot be differentiated based on their objective ``physical'' characteristics, how can we form judgements regarding their nature?

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality (a process henceforth referred to as \emph{simulation monitoring}, Makowski, 2018; Makowski, Sperduti, et al., 2019). With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy (Michael \& Sanson, 2021; Susmann et al., 2021), people tend to draw on peripheral contextual cues (\textbf{Figure 1}), such as the source of the stimulus (e.g., which journal was the information published in), and its credibility, authority and expertise, to help facilitate their evaluation (Michael \& Sanson, 2021; Petty \& Cacioppo, 1986; Susmann et al., 2021). However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are often mass-shared with little context) makes this task progressively difficult (Berghel, 2018; Y. Chen et al., 2015). Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial role in simulation monitoring, with factors such as cognitive style, prior beliefs, and personality traits (Bryanov \& Vziatysheva, 2021; Ecker et al., 2022; Sindermann et al., 2020). For instance, individuals with stronger analytical reasoning skills have been found to better discriminate real from fake stimuli (Pehlivanoglu et al., 2021; Pennycook \& Rand, 2019), and prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information that is in line with one's expectations (Britt et al., 2019). Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news (Piksa et al., 2022; Sindermann et al., 2020).

@@

Another issue is the impact on reality judgements of the prior explicit instruction that ``about half of the faces were AI-generated and the other half real photos''. Given this prior information given to participants, it might seems like our enthusiasm pertaining to the finding that most people did indeed believe a high number of stimuli to be fake might be unwarranted, since it simply affirms participants followed the instructions. However, even if that was the case, the finding that our beliefs of reality can be so easily re-programmed with simple instructions and lead to high-confidence answers remains an interesting phenomenon. Moreover, it is to note that the paradigm did not instruct participants to balance their answers according to a certain distribution (e.g., 50-50), merely providing them a description of the dataset. The fact that no presentation order effect was found on reality beliefs suggests that participants did not try to actively distribute their responses to match the instructions, in which case we would have expected a different pattern, for instance the first few items judged as real (the initial ``true'' belief of the participants), and progressively a hard bias towards responding ``fake'' (as participants realize that all stimuli are of similar nature and that they have to ``make up'' for the prevalence of their ``real'' answers to fulfill the expected proportion of responses given the instructions).

That said, the potential demand effect of the instructions still exists, and a control condition without the cover story with AI-generated images would in-principle be able to mitigate such confounds to some extent. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum (simply instructing them to discriminate real from fake without providing some background information regarding the context and defining what is meant by ``fake'' seems hardly feasible). That being the case, future studies should study the impact of these higher-order expectations on ratings (for instance, Tucciarelli et al., 2020 found that merely mentioning that some faces were AI generated decreased, on average, the trustworthiness ratings for all faces) as well as on the simulation monitoring process itself (i.e., the ``criterion'': would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., ``most of the images \emph{but a few} are real'' vs.~``most of the images \emph{but a few} are fake'') or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs.~real to participants.

@@

\leavevmode\vadjust pre{\hypertarget{ref-makowski2019phenomenal}{}}%
Makowski, D., Sperduti, M., Pelletier, J., Blondé, P., La Corte, V., Arcangeli, M., Zalla, T., Lemaire, S., Dokic, J., Nicolas, S.others. (2019). Phenomenal, bodily and brain correlates of fictional reappraisal as an implicit emotion regulation strategy. \emph{Cognitive, Affective, \& Behavioral Neuroscience}, \emph{19}(4), 877--897.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rhodes2006evolutionary}{}}%
Rhodes, G.others. (2006). The evolutionary psychology of facial beauty. \emph{Annual Review of Psychology}, \emph{57}, 199.
